\documentclass[main.tex]{subfiles}

\begin{document}
\pagestyle{plain}
\setcounter{chapter}{0}

\chapter{Basic Setting}
\label{chap:basic-setting}

\section{The Abelian Sandpile Model}

The Abelian sandpile model is a cellular automaton introduced by Bak, Tang and Wiesenfeld. A $d$-dimensional lattice holds a non-negative integer $h(x)$ at each site $x$, the number of grains. A site is \emph{stable} when $h(x) < T$, where $T$ is the toppling threshold; it is \emph{unstable} when $h(x) \ge T$. An unstable site \emph{topples}: it loses $T$ grains and sends one grain to each of its $2d$ nearest neighbours. Grains that leave the boundary are lost (open boundary condition). Repeated toppling eventually returns every site to a stable configuration; this final configuration is independent of the order of topplings (the Abelian property).

In two dimensions the grid is $N \times N$, the threshold is $T=4$, and each site has four orthogonal neighbours. In three dimensions the cube is $n \times n \times n$, the threshold is $T=6$, and each site has six neighbours ($\pm x,\pm y,\pm z$).

\section{Implementation of the Sandpile}
\label{sec:impl-sandpile}

The model is implemented in Python with NumPy. Two classes, \texttt{Sandpile2D} and \texttt{Sandpile3D}, share the same structure; we describe the 2D case for brevity. The grid is stored as a NumPy integer array. Stabilisation does not rescan the whole grid each step; instead it keeps a queue of currently unstable sites, so the running time is $O(N^2 + \text{number of topples})$. The core routine is:

\begin{lstlisting}[language=Python, caption={Stabilisation by queue (2D, simplified).}]
def stabilize(grid, N, T=4):
    queue = deque(sites where grid >= T)
    topples = 0
    while queue:
        (x, y) = queue.popleft()
        if grid[x, y] < T:        # already stabilised by a neighbour
            continue
        grid[x, y] -= T           # topple
        topples += 1
        for (nx, ny) in neighbours(x, y):   # up, down, left, right
            if inside(nx, ny, N):
                grid[nx, ny] += 1
                if grid[nx, ny] >= T:
                    queue.append((nx, ny))
    return grid, topples
\end{lstlisting}

Each toppling event removes $T$ grains from a site and distributes one grain to every in-bounds neighbour; any grain crossing the boundary is discarded. The 3D version is identical except that the threshold is $6$ and each toppling distributes to six neighbours instead of four.

\section{Experiment Design}
\label{sec:experiment-design}

The experiments follow the instructions given in the email correspondence that initiated this project. The protocol for a single trial is:

\begin{enumerate}
    \item Start with an $N \times N$ (2D) or $n \times n \times n$ (3D) grid.
    \item Put a fixed \emph{background} of grains on every site: $3$ grains in 2D, $5$ grains in 3D. The background is chosen one below the threshold, so the grid is stable but on the verge of toppling.
    \item Select three random distinct sites and add one extra grain to each. At least one site now reaches the threshold and an avalanche begins.
    \item Stabilise the grid with the routine of Section~\ref{sec:impl-sandpile}.
    \item Save the final stable grid to disk and record statistics: total topplings, number of distinct toppled sites, the grain distribution, the seed, and the perturbed positions.
\end{enumerate}

This trial is repeated many times ($10\,000$ in 2D, $2\,000$ in 3D). Each trial uses an independent random seed $s_i = s_0 + i$, so the perturbation sites differ every trial while the experiment as a whole remains reproducible. A central configuration file (\texttt{config.py}) holds all parameters---grid size, threshold, background, number of perturbation sites, number of trials, seed and output directories---so that the whole experiment can be retuned from one place. The driver (\texttt{main.py}) reads these parameters, runs the batch, and saves grids and aggregate statistics incrementally so that progress survives an interruption.

After the batch completes, the same driver produces the classification study of Chapter~\ref{chap:classification}: for each direction $(n,m)$ with $0 \le n,m \le 7$ it extracts the height pattern far from the centre along that direction and arranges the resulting images in a table.

\section{Visualisation}
\label{sec:visualisation}

The module \texttt{visualize.py} (built on Matplotlib) provides interactive viewers for exploring the saved grids. Each grain count is mapped to a fixed colour via a discrete 7-colour colormap (off-white~$\rightarrow$~yellow~$\rightarrow$~blue~$\rightarrow$~green~$\rightarrow$~orange~$\rightarrow$~brown~$\rightarrow$~red), so that the height pattern is read directly as a coloured image.

\subsection{2D Interactive Viewer (\texttt{-{}-view2d})}

The 2D viewer opens a Matplotlib window with a trial slider to browse grid snapshots. Red $\bigstar$ markers indicate the three perturbation sites for each trial. The title bar displays the trial index, total topple count, and distinct-toppled-site count. Keyboard shortcuts $\leftarrow$~/~$\rightarrow$ (or A~/~D) navigate between trials.

\subsubsection{Patch Inspection Feature}
The 2D viewer includes a local inspection tool: hovering the mouse over any grid cell shows a live \textbf{$18\times18$ preview inset} (top-left corner of the window) displaying the local neighbourhood around the cursor. Left-clicking or pressing the \texttt{S} key \textbf{saves} the patch as an annotated PNG image. In the saved image every cell is labelled with its exact grain count (black text for $\le 2$ grains, white for $\ge 3$), making it possible to read precise height values without squinting at the colour map. Saved patches are written to \texttt{outputs/2d/local\_patches/} with filenames of the form \texttt{patch\_trial\_NNNNNN\_xX\_yY.png}.

This feature is the primary tool for examining local periodic colour structures along directional lines (see Chapter~\ref{chap:classification}), inspecting grain-count patterns at intersection points, and capturing exact configurations for inclusion in this report.

\subsection{3D Interactive Viewer (\texttt{-{}-view3d})}

The 3D viewer provides two orthogonal sliders---trial index and $z$-slice depth---for browsing 3D grids one 2D layer at a time. The grid is cached per trial, so switching $z$ within the same trial is instantaneous. Keyboard shortcuts: $\leftarrow$~/~$\rightarrow$ navigate $z$-slices, $\uparrow$~/~$\downarrow$ (or W~/~S) navigate trials.

These interactive tools were used both to spot the regularities reported in Chapter~\ref{chap:classification} and to produce the figures in this report.

\end{document}
